\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-21 papers/jss/article.Rnw
Mostly work on section 2 to address some flaws identified by Jeroen, Karl, and others.
Move up the basic description of the protocol buffer schema from
section 3 to section 2, including the example of how protocol buffers
are manipulated with this package.

Revert a regression -- fix the reference to BSON and MessagePack by
putting the citations next to the text about the R interfaces rather
than the formats themselves (re-apply fix from Dirk).

Add an explicit transition from section 2 to section 3 as the last sentence of 2.

Define payload at the beginning of section 3 just once, so we don't

A number of binary formats based on \texttt{JSON} have been proposed
that reduce the parsing cost and improve efficiency.  \pkg{MessagePack}
\citep{msgpackR} and \pkg{BSON} \citep{rmongodb} both have R
interfaces, but these formats lack a separate schema for the seralized
data and thus still duplicate field names with each message sent over
the network or stored in a file.  Such formats also lack support for
versioning when data storage needs evolve over time, or when
application logic and requirement changes dictate updates to the
message format.

@@

% JO: I'm not sure where to put this paragraph. I think it is too technical
% for the introduction section. Maybe start this section with some explanation
% of what a schema is and then continue with showing how PB implement this?
Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Protocol Buffers \citep{protobuf}, Apache Thrift, and Apache Avro provide a compact
well-documented schema for cross-language data structures and
efficient binary interchange formats.
Since the schema is provided separately from the encoded data, the data can be
efficiently encoded to minimize storage costs of the stored data when compared with simple
``schema-less'' binary interchange formats. 
The schema can be used to generate classes for statically-typed programming languages
such as C++ and Java, or can be used with reflection for dynamically-typed programming
languages.

@@

%% TODO(de,ms)  What follows is oooooold and was lifted from the webpage
%%              Rewrite?
Protocol Buffers can be described as a modern, language-neutral, platform-neutral,
extensible mechanism for sharing and storing structured data.  Since their
introduction, Protocol Buffers have been widely adopted in industry with
applications as varied as %database-internal messaging (Drizzle), % DE: citation?
Sony Playstations, Twitter, Google Search, Hadoop, and Open Street Map.  
% TODO(DE): This either needs a citation, or remove the name drop
% MS: These are mostly from blog posts, I can't find a good reference
% that has a long list, and the name and year citation style seems
% less conducive to long lists of marginal citations like blog posts
% compared to say concise CS/math style citations [3,4,5,6]. Thoughts?

While traditional IDLs have at times been criticized for code bloat and
complexity, Protocol Buffers are based on a simple list and records
model that is compartively flexible and simple to use.

Some of the key features provided by Protocol Buffers for data analysis
include:

@@

@@

@@

%While benchmarks are not available, Google states on the project page that in
%comparison to XML, protocol buffers are at the same time \textsl{simpler},
%between three to ten times \textsl{smaller}, between twenty and one hundred
%times \textsl{faster}, as well as less ambiguous and easier to program.

%The flexibility of the reflection-based API is particularly well
%suited for interactive data analysis.

% XXX Design tradeoffs: reflection vs proto compiler

For added speed and efficiency, the C++, Java, and Python bindings to
Protocol Buffers are used with a compiler that translates a Protocol
Buffer schema description file (ending in \texttt{.proto}) into
language-specific classes that can be used to create, read, write and
manipulate Protocol Buffer messages.  The R interface, in contrast,
uses a reflection-based API that is particularly well-suited for
interactive data analysis.  
All messages in R have a single class
structure, but different accessor methods are created at runtime based
on the named fields of the specified message type.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{protobuf-distributed-system-crop.pdf}
\end{center}
\caption{Example protobuf usage}
\label{fig:protobuf-distributed-usecase}
\end{figure}

@@

This section describes how to use the R API to create and manipulate
protocol buffer messages in R, and how to read and write the
binary \emph{payload} of the messages to files and arbitrary binary
R connections.

The two fundamental building blocks of Protocol Buffers are \emph{Messages}
and \emph{Descriptors}.  Messages provide a common abstract encapsulation of
structured data fields of the type specified in a Message Descriptor.
Message Descriptors are defined in \texttt{.proto} files and define a
schema for a particular named class of messages.

Table~\ref{tab:proto} shows an example \texttt{.proto} file which
defines the \texttt{tutorial.Person} type.  The R code in the right
column shows an example of creating a new message of this type and
populating its fields.  A \texttt{.proto} file may contain one or more
message types, and each message type has one or more fields.  A field
is specified with a unique number, a name, a value type, and a field
rule specifying whether the field is optional, required, or repeated.
The supported value types are numbers, enumerations, booleans,
strings, raw bytes, or other nested message types.
The \texttt{.proto} file syntax for defining the structure of protocol
buffer data is described comprehensively on Google Code\footnote{See 
\url{http://code.google.com/apis/protocolbuffers/docs/proto.html}.}.

@@

The binary representation of the message (often called the payload)
does not contain information that can be used to dynamically
infer the message type, so we have to provide this information
to the \texttt{read} function in the form of a descriptor :

@@

The previous sections discussed functionality in the \pkg{RProtoBuf} package
for creating, manipulating, parsing and serializing Protocol Buffer
messages of a specific pre-defined schema.  The package also provides
methods for converting arbitrary R data structures into protocol
buffers and vice versa with a universal R object schema. The \texttt{serialize\_pb} and \texttt{unserialize\_pb}
functions serialize arbitrary R objects into a universal Protocol Buffer 
message:

>>>>

A number of binary formats based on \texttt{JSON} have been proposed
that reduce the parsing cost and improve efficiency.  \pkg{MessagePack}
and \pkg{BSON} both have R
interfaces \citep{msgpackR,rmongodb}, but these formats lack a separate schema for the serialized
data and thus still duplicate field names with each message sent over
the network or stored in a file.  Such formats also lack support for
versioning when data storage needs evolve over time, or when
application logic and requirement changes dictate updates to the
message format.

@@

% JO: I'm not sure where to put this paragraph. I think it is too technical
% for the introduction section. Maybe start this section with some explanation
% of what a schema is and then continue with showing how PB implement this?
% MS: Yes I agree, tried to address below.

@@

%% TODO(de,ms)  What follows is oooooold and was lifted from the webpage
%%              Rewrite?
Protocol Buffers are a modern, language-neutral, platform-neutral,
extensible mechanism for sharing and storing structured data.  Some of
the key features provided by Protocol Buffers for data analysis include:

@@

% Lets place this at the top of the page or the bottom, or on a float
% page, but not just here in the middle of the page.
\begin{figure}[tbp]
\begin{center}
\includegraphics[width=\textwidth]{protobuf-distributed-system-crop.pdf}
\end{center}
\caption{Example protobuf usage}
\label{fig:protobuf-distributed-usecase}
\end{figure}

@@

While traditional IDLs have at times been criticized for code bloat and
complexity, Protocol Buffers are based on a simple list and records
model that is flexible and simple to use.  The schema for structured
protocol buffer data is defined in \texttt{.proto} files which may
contain one or more message types.  Each message type has one or more
fields.  A field is specified with a unique number, a name, a value
type, and a field rule specifying whether the field is optional,
required, or repeated.  The supported value types are numbers,
enumerations, booleans, strings, raw bytes, or other nested message
types.  The \texttt{.proto} file syntax for defining the structure of protocol
buffer data is described comprehensively on Google Code\footnote{See 
\url{http://code.google.com/apis/protocolbuffers/docs/proto.html}.}.
Table~\ref{tab:proto} shows an example \texttt{.proto} file which
defines the \texttt{tutorial.Person} type.  The R code in the right
column shows an example of creating a new message of this type and
populating its fields.

% The schema can be used to generate model classes for statically-typed programming languages
%such as C++ and Java, or can be used with reflection for dynamically-typed programming
%languages.

% TODO(mstokely): Maybe find a place to add this?  
% Since their
% introduction, Protocol Buffers have been widely adopted in industry with
% applications as varied as %database-internal messaging (Drizzle), % DE: citation?
% Sony Playstations, Twitter, Google Search, Hadoop, and Open Street
% Map.

% TODO(DE): This either needs a citation, or remove the name drop
% MS: These are mostly from blog posts, I can't find a good reference
% that has a long list, and the name and year citation style seems
% less conducive to long lists of marginal citations like blog posts
% compared to say concise CS/math style citations [3,4,5,6]. Thoughts?

% The schema can be used to generate classes for statically-typed programming languages
% such as C++ and Java, or can be used with reflection for dynamically-typed programming
% languages.

@@

For added speed and efficiency, the C++, Java, and Python bindings to
Protocol Buffers are used with a compiler that translates a Protocol
Buffer schema description file (ending in \texttt{.proto}) into
language-specific classes that can be used to create, read, write and
manipulate Protocol Buffer messages.  The R interface, in contrast,
uses a reflection-based API that is particularly well-suited for
interactive data analysis.  
All messages in R have a single class
structure, but different accessor methods are created at runtime based
on the named fields of the specified message type, as described in the
next section.

@@

@@

This section describes how to use the R API to create and manipulate
protocol buffer messages in R, and how to read and write the
binary representation of the message (often called the \emph{payload}) to files and arbitrary binary
R connections.
The two fundamental building blocks of Protocol Buffers are \emph{Messages}
and \emph{Descriptors}.  Messages provide a common abstract encapsulation of
structured data fields of the type specified in a Message Descriptor.
Message Descriptors are defined in \texttt{.proto} files and define a
schema for a particular named class of messages.

@@

The binary representation of the message
does not contain information that can be used to dynamically
infer the message type, so we have to provide this information
to the \texttt{read} function in the form of a descriptor :

@@

The previous sections discussed functionality in the \pkg{RProtoBuf} package
for creating, manipulating, parsing and serializing Protocol Buffer
messages of a pre-defined schema.  This is useful when there are
pre-existing systems with defined schemas or significant software
components written in other languages that need to be accessed from
within R.

The package also provides methods for converting arbitrary R data structures into protocol
buffers and vice versa with a universal R object schema. The \texttt{serialize\_pb} and \texttt{unserialize\_pb}
functions serialize arbitrary R objects into a universal Protocol Buffer 
message:
